Safety-aligned large language models (LLMs) refuse harmful requests, and a
jailbreak is any input that defeats that refusal. This work, named prayoga after
the Sanskrit term for applied practice, asks whether jailbreak and prompt
injection, hypnotic suggestion, and the tantric act of vaś\=\i kara\d{n}a
(subjugation) share one mechanism. That mechanism is the capture of a system's
output policy by an injected context that suppresses a monitoring faculty while
co-opting automatic generation. The thesis is split into three labelled
tiers, namely mechanism, functional analogy, and metaphor with a falsifiable
core, and tested by mechanistic interpretability on open-weight chat models
(Gemma-2-2B, Gemma-2-9B, and Qwen2.5-3B) and by behavioural probing of a frontier
model (Claude Opus 4.8, accessed through the \texttt{claude -p} interface). On the
AgentDojo agentic benchmark this model sustains $0.80$ task utility at a $0.00$
attack-success rate, which fixes the resilient reference end. The
empirical hinge is an order parameter for refusal: the normalized projection of
the residual stream onto a difference-in-means refusal direction. This quantity is
invariant across paraphrase orbits (a between-class to within-orbit $F$-ratio of
$19.2$ versus $0.65$ for a random direction in Gemma) and collapses under injection
(by $34\%$ in Gemma and $45\%$ in Qwen). The statement that refusal is a symmetry
and a jailbreak is symmetry-breaking is therefore a measurement, not a metaphor.
The same direction is causal and dosable. Ablation raises the attack success rate
from $0.00$ to $0.90$, addition raises over-refusal from $0.05$ to $1.00$, and a
four-parameter logistic gives a half-maximal ablation strength of EC50 $=0.33$
($R^2=0.996$). A calibrated coefficient sweep then shows single-direction addition
is sufficient in both families, so an apparent cross-family asymmetry was a dose
artifact rather than a structural difference. A cross-axis triangulation supplies
the sharpest bound. Injection collapses the internal monitor readouts, yet a
content-faithful judge finds genuine behavioural capture in only $1$ to $2.4\%$ of
cases. Internal collapse is therefore necessary but not sufficient for capture, and
the unification holds at the representational level alone. Falsification is treated
as a starting point rather than an endpoint. Each falsified metaphysical claim
triggers a designed follow-up that recovers the real signal, including a
mid-network truth-evaluation direction that transfers across topics at $0.96$ above
a chance-level surface baseline. The most speculative avasth\=atraya and tur\=\i ya
claims do not survive surface and anisotropy controls. The contribution is a map of
where a cross-domain symmetry is measured, where it is analogy, and where it breaks.
