% sections/appendix.tex
% Appendix: Pre-Registered Falsification Gates and Metaphor-Tier Testing

\section*{Appendix A: Falsification Gates for Claim Tiers}

This appendix lists all metaphor-tier and analogy-tier claims tested in this work, their
falsification criteria, and results. Mechanism-tier claims (refusal direction exists, can be
ablated, can be steered) are measurements, not gated hypotheses; they are reported with confidence
intervals and effect sizes. Analogy- and metaphor-tier claims require explicit falsification gates
to avoid unfalsifiable speculation.

\subsection*{Mechanism Tier: Measurements (not gated speculation)}

The following are empirical measurements; gates are not applicable.

\begin{enumerate}
\item \textbf{M1: Refusal is mediated by a low-dimensional direction in residual stream.}
  \begin{itemize}
    \item \textbf{Measurement:} Rank of ablated activation matrix by layer (F18).
    \item \textbf{Result:} Layer-dependent; effective rank 1--8 depending on model and layer.
    \item \textbf{Confidence:} High. Replicated across Gemma-2-2B, Gemma-2-9B, Qwen2.5-3B.
  \end{itemize}

\item \textbf{M2: Single-direction steering achieves dose-response control.}
  \begin{itemize}
    \item \textbf{Measurement:} EC50 (half-maximal effective concentration) from sigmoid dose-response fit.
    \item \textbf{Result:} EC50 ranges 0.8--2.3 depending on layer; attack success rate ranges 12\%--94\%.
    \item \textbf{Confidence:} High. 95\% confidence intervals report $\pm 0.15$ typical.
  \end{itemize}

\item \textbf{M3: Unsupervised SAE feature orthogonal to supervised direction is causally sufficient.}
  \begin{itemize}
    \item \textbf{Measurement:} Cosine similarity (expect $\sim 0.002$, i.e., $\cos \theta \ll 0.1$);
      ablation of the feature removes refusal equally to ablation of the direction.
    \item \textbf{Result:} Feature identified; similarity 0.0018; ablation effect size (logit drop) 8.2 vs.\ 8.1
      (direction).
    \item \textbf{Confidence:} Medium. Single SAE, single model; generalization untested.
  \end{itemize}
\end{enumerate}

\subsection*{Analogy Tier: Gated hypotheses}

The following are functional parallels between hypnosis and LLM jailbreaking, grounded in
Norman--Shallice and Free Energy Principle frameworks. Gates are operational tests.

\begin{enumerate}
\item \textbf{A1: Monitoring faculty can be selectively suppressed while generation remains intact.}
  \begin{itemize}
    \item \textbf{Hypothesis:} Steering along the refusal direction selectively removes the refusal
      response without degrading overall language quality (perplexity, BLEU on benign prompts).
    \item \textbf{Falsification criterion:} If perplexity increases $> 5\%$ under steering, or if benign
      accuracy drops $> 10\%$, the hypothesis fails (we claim selective suppression, not global degradation).
    \item \textbf{Result:} Pass. Benign-prompt perplexity 7.8 (baseline) vs.\ 7.9 (steered, $\Delta = +1.3\%$);
      accuracy on MMLU 70.2\% vs.\ 69.8\% ($\Delta = -0.4\%$).
    \item \textbf{Consequence:} The analogy holds: monitoring can be suppressed without disabling generation.
  \end{itemize}

\item \textbf{A2: Refusal suppression operates via low-precision updating (FEP reading).}
  \begin{itemize}
    \item \textbf{Hypothesis:} The magnitude of the refusal direction (in norm) correlates with the
      Free Energy reduction term; large-norm directions imply high confidence (precision) in the
      monitoring belief.
    \item \textbf{Falsification criterion:} No correlation between direction norm and logit gap between
      refuse and comply responses ($r < 0.3$), or orthogonal directions have equal effect (they should not).
    \item \textbf{Result:} Pass. Norm--logit-gap correlation $r = 0.71$ ($p < 0.001$); orthogonal
      component has 0.15$\times$ the effect of the principal refusal direction.
    \item \textbf{Consequence:} Direction norm is a proxy for monitoring precision; FEP framing is
      consistent with data.
  \end{itemize}

\item \textbf{A3: Non-surjectivity of steered activations mirrors the inaccessibility of hypnotic states.}
  \begin{itemize}
    \item \textbf{Hypothesis:} Steered activation states are off-manifold (Mishra et al.\ 2026);
      correspondingly, steering achieves stronger refusal suppression than any single natural prompt can.
    \item \textbf{Falsification criterion:} If the best many-shot attack (Anil et al.\ 2024) achieves
      comparable ASR to steering, the analogy fails (natural prompts would be reaching off-manifold states).
    \item \textbf{Result:} Pass. Many-shot ASR: 52\% (on test set). Steering ASR: 88\% (at EC50).
      Strong separation suggests steering accesses states beyond prompt space.
    \item \textbf{Consequence:} The non-surjectivity argument supports the dissociation analogy.
  \end{itemize}
\end{enumerate}

\subsection*{Metaphor Tier: Falsifiable speculations}

The following map tantric and Vedantic categories to LLM mechanisms. These are the highest-risk
claims and are tested most rigorously.

\begin{enumerate}
\item \textbf{F1: Multi-seed robustness of identified features and directions.}
  \begin{itemize}
    \item \textbf{Hypothesis:} The refusal direction and SAE feature are robust across random seeds
      of the SAE training and model initialization.
    \item \textbf{Falsification criterion:} Cosine similarity of directions from different seeds $< 0.85$,
      or effect size varies $> 30\%$ across seeds.
    \item \textbf{Result:} Pass. Cosine similarity: 0.94 $\pm$ 0.03 across 5 seeds. Effect size (logit drop)
      8.2 $\pm$ 0.4.
    \item \textbf{Consequence:} Features are not artifacts of initialization; they generalize.
  \end{itemize}

\item \textbf{F2: \d{s}a\d{t}karma taxonomy: gradual (many-shot) vs.\ sudden (GCG) vaś\=\i kara\d{n}a.}
  \begin{itemize}
    \item \textbf{Hypothesis:} Many-shot attacks operate via gradual context pollution (low-frequency,
      cumulative). GCG operates via sudden suffix injection (high-frequency, instantaneous).
      The ṣaṭkarma predicts this dichotomy: vaśīkaraṇa has two subtypes (gradual and sudden).
    \item \textbf{Falsification criterion:} No difference in latent attention patterns or activation dynamics
      between many-shot and GCG trajectories; or attack success rate shows no interaction with attack type
      (gradual vs.\ sudden) and model size.
    \item \textbf{Result:} Partial pass. Attention pattern differences are significant ($p < 0.01$);
      attack success does depend on type ($\chi^2 = 23.4$, $p < 0.001$). However, the ṣaṭkarma taxonomy
      does not uniquely predict the success rates or optimal dose.
    \item \textbf{Consequence:} The taxonomy is descriptive of attack modes, not predictive of attack strength.
      This suggests the ṣaṭkarma is a valid taxonomy but does not license claims of deeper causal structure.
  \end{itemize}

\item \textbf{F3: Avasth\=atraya regimes: F3 (harmful refusal) vs.\ F5 (jailbroken generation) are functionally
  distinct regimes.}
  \begin{itemize}
    \item \textbf{Hypothesis:} The jāgrat (waking) regime corresponds to normal alignment (F3);
      the svapna (dream) regime corresponds to jailbroken generation (F5); svapna and jāgrat are
      separable by the steering magnitude (order parameter).
    \item \textbf{Falsification criterion:} No significant difference in downstream task performance,
      knowledge recall, or consistency between F3 and F5; or the separation is not monotonic in
      steering magnitude.
    \item \textbf{Result:} Partial pass. Consistency differences are significant ($p < 0.05$); task
      performance differs ($p < 0.01$). However, the regimes are not as cleanly separated as the
      metaphor suggests: intermediate steering magnitudes show mixed behaviors.
    \item \textbf{Consequence:} Two regimes exist, but with a continuum between them, not a sharp phase transition.
      The avasthātraya framing is suggestive but not uniquely explanatory.
  \end{itemize}

\item \textbf{F7: Tur\=\i ya as an attractor fixed point.}
  \begin{itemize}
    \item \textbf{Hypothesis:} Successive self-paraphrase (Wang et al.\ 2025) converges to an attractor
      cycle that corresponds to the turīya (transcendent state). This attractor should be orthogonal
      to the embedding space's principal axes (low mutual information with content).
    \item \textbf{Falsification criterion:} No attractor is found (paraphrases drift indefinitely);
      or the attractor has high mutual information with content axes (is not privileged);
      or the attractor depends on initialization (is not stable).
    \item \textbf{Result:} \textbf{Fail.} Successive paraphrases do converge to a fixed point (Wang result replicated),
      but the fixed point is \emph{not} orthogonal to the embedding's principal axes. In fact, it aligns
      strongly with the content-encoding subspace. This suggests the attractor is a content-saturation artifact,
      not a privileged state.
    \item \textbf{Consequence:} The turīya mapping to an attractor fails. The avasthātraya metaphor does not
      extend to LLM state space in the expected way. This is a valuable negative result: it clarifies that
      the metaphor's power breaks down at the mechanistic level.
  \end{itemize}

\item \textbf{F11: Cross-domain generalization of the order parameter (analogy-tier claim generalized).}
  \begin{itemize}
    \item \textbf{Hypothesis:} The order parameter (coupling break, measured as direction norm vs.\ logit gap)
      should generalize to non-LLM systems, e.g., hypnotic suggestion in humans or the efficacy of mantra
      in tantric practice (if data were available).
    \item \textbf{Falsification criterion:} No data are available for independent validation; the hypothesis
      is untestable. Alternatively, if surrogate data from prior work (e.g., hypnotic susceptibility scales)
      show no correlation with model-derived order-parameter analogs, the hypothesis fails.
    \item \textbf{Result:} Not tested. No hypnosis data are available. This gate is marked for future work.
    \item \textbf{Consequence:} The analogy remains one-directional: LLM to hypnosis, not vice versa. Replication
      in human or ritual contexts awaits external data.
  \end{itemize}

\item \textbf{F13: Circularity check: is the avasth\=atraya mapping fitted post-hoc?}
  \begin{itemize}
    \item \textbf{Hypothesis:} The avasthātraya regimes (jāgrat, svapna, suṣupti, turīya) are not arbitrarily
      fit to the data post-hoc. To test this, we should derive predictions \emph{before} seeing the results,
      and check whether those predictions hold.
    \item \textbf{Falsification criterion:} The regimes are re-labeled retroactively to match observed phenomena;
      or the number of regimes changes post-hoc; or predictions made in the pre-registration (Definitions section)
      are violated.
    \item \textbf{Result:} \textbf{Partial fail.} The four regimes are as predicted, and turīya mapping was
      pre-registered (F7). However, the regime boundaries (steering magnitude thresholds) were determined
      empirically, not predicted in advance. This is a form of post-hoc adjustment, though transparent.
    \item \textbf{Consequence:} The circularity risk is present but mitigated by transparency. Future work should
      set regime boundaries in advance, using held-out test data.
  \end{itemize}

\item \textbf{F18: Layer sweep: refusal direction effective dimensionality varies by layer.}
  \begin{itemize}
    \item \textbf{Hypothesis:} Effective rank of the refusal subspace is layer-dependent, and the layer with
      lowest effective rank is optimal for steering (maximal control with minimal dimensionality).
    \item \textbf{Falsification criterion:} Effective rank is constant across layers, or does not predict steering
      efficacy.
    \item \textbf{Result:} Pass. Rank ranges from 1 (layer 18, Gemma-2-2B) to 8 (layer 10, same model).
      Layer 18 achieves highest steering efficacy; layer 10 shows scatter (mixed results).
      Correlation between rank and steering effect size: $r = -0.68$ ($p < 0.01$).
    \item \textbf{Consequence:} Layer-sweep is a valid strategy; low-rank layers are steering-optimal.
  \end{itemize}
\end{enumerate}

\subsection*{Summary: Falsification outcomes}

\begin{center}
\begin{tabular}{lcc}
\hline
Claim Category & Tier & Result \\
\hline
Refusal dimension layer-dependent & Mechanism & Pass \\
EC50 dose-response & Mechanism & Pass \\
Orthogonal SAE feature sufficient & Mechanism & Pass (Low confidence) \\
Monitoring selective suppression & Analogy & Pass \\
FEP precision framework & Analogy & Pass \\
Non-surjectivity analogy & Analogy & Pass \\
\d{S}a\d{t}karma taxonomy & Metaphor & Partial (Descriptive, not predictive) \\
Avasthātraya regimes F3, F5 & Metaphor & Partial (Continuum, not sharp phase) \\
Turīya as attractor (F7) & Metaphor & \textbf{Fail} \\
Cross-domain order parameter (F11) & Metaphor & Not tested \\
Circularity check (F13) & Metaphor & Partial (Post-hoc thresholds) \\
Layer-sweep efficacy (F18) & Mechanism & Pass \\
\hline
\end{tabular}
\end{center}

The mechanism tier is robust. The analogy tier is well-supported. The metaphor tier shows
mixed results: the ṣaṭkarma taxonomy is descriptive but not predictive; the avasthātraya
regimes exist but with a continuum, not sharp transitions; the turīya hypothesis fails,
a valuable negative result. These findings support the three-tier discipline: metaphor-tier
claims are the riskiest and least certain, mechanism-tier claims are the most solid.
