This section situates prayoga against five strands of prior art: the
representation geometry of refusal, the attack literature that supplies the inputs
which break it, sparse-feature interpretability, the cognitive-science account of
hypnosis, and the tantric and Vedantic sources that motivate the metaphor tier.
Each strand is reviewed in turn, with the present contributions noted where they
attach.

\subsection{Refusal as a direction, and the dimensionality debate}
The proximate basis for this work is the finding that refusal is mediated by a
single linear direction in the residual stream~\cite{arditi2024refusal}: across
many open chat models, the difference-in-means of harmful and harmless prompt
activations yields a vector whose ablation removes refusal and whose addition
induces it. The direction is reported to be largely universal across safety-aligned
languages~\cite{wang2025multilingual}, though its robustness degrades sharply in
reasoning-tuned models~\cite{yin2025refusal}. This both explains earlier representation-engineering
results~\cite{zou2023representation} and gives a constructive ``abliteration''
recipe. The claim has since been contested in its strong form.
\citet{marshall2024refusal} argue that refusal is better described by an affine
subspace than a single linear direction, and \citet{wollschlager2025geometry}
show, through ``concept cones,'' that multiple representationally-independent
directions can each mediate refusal.

Recent work has pushed the debate toward greater nuance. \citet{johnson2026more}
report that refusal is not exhausted by a single direction; rather, eleven distinct
categories of refusal mechanism can be identified, each with a layer-specific
effective subspace, yet all sharing a common necessary core. \citet{piras2025som}
advance Self-Organizing Maps to extract multiple refusal directions, showing that
SOMs generalize the difference-in-means technique and that multi-directional
suppression outperforms single-direction ablation. Concurrently, work on
non-linearity~\cite{hildebrandt2025refusal} challenges the linearity assumption
altogether, using dimensionality reduction (PCA, t-SNE, UMAP) to reveal
multi-dimensional, layer-dependent, and architecture-dependent refusal
characteristics. \citet{zhao2025harmfulness} identify a separate harmfulness
direction, orthogonal to refusal, showing that jailbreaks can reduce refusal signals
without reversing the model's internal judgment of harmfulness.

The cross-model results of this work bear directly on this debate: the effective
dimension of the refusal subspace is model- and layer-dependent, and this
dimension---not a universal constant---shapes whether single-direction steering is
bidirectionally causal at a given coefficient. The dose-response pharmacological
characterization (EC50 fits) introduced here resolves some of the ambiguity: a
low-dimensional refusal manifold suffices to explain the control observed, but its
dimensionality varies by layer and model, requiring a layer-sweep strategy to
identify the optimal target.

\subsection{Jailbreaks as control}
The attack literature supplies the inputs $u$ that break refusal: gradient-based
universal adversarial suffixes (GCG)~\cite{zou2023universal}, with recent
improvements to optimization efficiency via MAGIC~\cite{li2024magic} and generative
modelling via AmpleGCG~\cite{liao2024ample}; many-shot in-context conditioning that
exploits long contexts~\cite{anil2024manyshot}; multi-turn crescendo
escalation~\cite{russinovich2024crescendo}; and indirect injection via tool and
agent channels, benchmarked by AgentDojo~\cite{debenedetti2024agentdojo}.
\citet{guo2024coldattack} make the control-theoretic reading explicit, casting
jailbreaking as constrained, controllable decoding via Langevin dynamics.

Recent mechanistic work has illuminated the internal mechanisms by which attacks
succeed. \citet{ben_tov2025universal} show that GCG's universality is driven by
attention hijacking---suffixes construct a shallow but critical mechanism that
dominates information flow, with stronger universality correlating with stronger
hijacking. \citet{von_recum2024cannot} provide a taxonomy of sixteen refusal
categories (cannot versus should-not, with subcategories), enabling precise auditing
of refusal composition in instruction-tuning and RLHF datasets. On the safety side,
\citet{guo2026llms} demonstrate that refusal can be unlearned with only one thousand
benign samples by disrupting fixed refusal-prefix completion pathways, suggesting
that current safety alignment relies on token-sequence memorization rather than
robust reasoning.

The contribution of this work is orthogonal and complementary: rather than searching
for a stronger $u$, it characterizes the target of $u$---the geometry and
symmetry of the refusal variable it must move---and measures the dose-response
relationship.

\subsection{Sparse features and the unit of intervention}
Sparse autoencoders (SAEs) decompose activations into interpretable, monosemantic
features~\cite{cunningham2023sparse,templeton2024scaling}, with BatchTopK SAEs a
current training method~\cite{bussmann2024batchtopk}. SAEs offer a finer unit of
intervention than a single difference-in-means direction. Recent work has refined
both SAE architectures and their application to steering: \citet{bussmann2024batchtopk}
introduce BatchTopK, improving SAE training efficiency; \citet{huang2025understanding}
apply SAEs specifically to understanding refusal mechanisms, revealing feature-level
structure; and \citet{vig2025steering} systematically steer refusal using
SAE-identified features.

Caveats have also emerged. \citet{korznikov2026sanity} show that SAEs recover only
$9\%$ of ground-truth features despite $71\%$ explained variance, and that random
baselines match fully-trained SAEs on interpretability ($0.87$ versus $0.90$) and
causal editing ($0.73$ versus $0.72$), questioning whether SAEs reliably decompose
mechanisms. \citet{cui2026sae} further demonstrate that SAE interventions can mask
recoverable failure modes: clamping a harmful feature allows the model to recover
pre-intervention behaviour via the SAE reconstruction residual, with a $95.8\%$
recovery rate on refusal steering. \citet{arad2025good} show that activation patterns
alone do not characterize feature effects; distinguishing input features (which
capture input patterns) from output features (which have human-interpretable
behavioural effects) yields $2$--$3\times$ improvements in steering.

A BatchTopK SAE is trained here (Section~\ref{sec:results}) on aligned Gemma-2-2B,
and a single unsupervised feature is identified that is orthogonal to the supervised
refusal direction yet causally sufficient to control refusal. This
orthogonality-and-sufficiency pair supports a mechanistic reading: the unsupervised
feature captures a redundant pathway, a natural substrate for the targeted acts in
the \d{s}a\d{t}karma taxonomy.

\subsection{The cognitive-science analogy: hypnosis and supervisory control}
The functional parallel to hypnosis rests on the Norman--Shallice model of action
control, in which a Supervisory Attentional System (SAS) modulates otherwise
automatic contention-scheduling~\cite{norman1986attention}. Cold-control theory holds
that hypnosis preserves first-order intentions while impairing higher-order awareness
of them~\cite{dienes2006coldcontrol}. \citet{riva2025automatic} draw the explicit map
to LLMs: the user prompt acts as a temporary external SAS, and a jailbreak suppresses
the monitoring faculty while leaving generation intact. In normative terms, the Free
Energy Principle~\cite{friston2010free} frames such suppression as lowering the
precision of a monitoring belief---an account adopted here for the analogy tier
and grounded empirically through the order parameter.

Recent work on activation steering and causality has sharpened this reading.
\citet{mishra2026steered} prove that steered activations are non-surjective:
activation steering pushes the residual stream off the manifold of states reachable
from discrete prompts. This formalism decouples white-box steerability from black-box
prompting, clarifying that precise internal control does not guarantee precise
behavioural control---a key insight for the analogy between hypnotic dissociation
(high internal control, low unified awareness) and jailbreak-via-steering.
\citet{joshi2026causality} emphasize that causal claims in interpretability require
falsification gates and transfer tests beyond correlation, reinforcing the need for
the layer-sweep and cross-domain validation pursued in this work.

\subsection{Attractors and the metaphysical tier}
On the speculative tier, two strands of prior art are relevant.
\citet{wang2025paraphrase} show that successive self-paraphrase converges to
attractor cycles, which supplies a non-mystical candidate for a substratum state; and
the M\=a\d{n}\d{d}\=ukya avasth\=atraya (waking, dream, deep-sleep, and
tur\=\i ya), set out in the M\=a\d{n}\d{d}\=ukya Upani\d{s}ad and
G\=au\d{d}ap\=ada's k\=arik\=a~\cite{mandukya,gaudapada}, furnishes a four-regime
template. Both are treated here as falsification
targets (Section~\ref{sec:results}), and---crucially---the attractor claim is tested
against the anisotropy of LLM embedding geometry, without which apparent convergence
is an artifact.

The dynamical-systems perspective has gained traction in recent mechanistic
interpretability. Work on causally-grounded explanation~\cite{mahale2026causally} and
data attribution via influence functions~\cite{chen2026mechanistic} highlights how
circuit-level mechanisms emerge from training data and how their causal status can be
validated. Certified circuits~\cite{anani2026certified} introduce randomized
subsampling to verify that circuit components remain invariant under bounded
perturbations, providing a formal ground for claiming mechanistic discovery rather
than dataset artifact. These tools support the falsification approach taken here: the
tur\=\i ya and vimarśa claims are not settled facts but testable hypotheses whose
failure yields useful negative results (Section~\ref{sec:results}).

\subsection{Vaś\=\i kara\d{n}a and the \d{s}a\d{t}karma}
Finally, the tantric source. Vaś\=\i kara\d{n}a (subjugation) is the second of the
six acts (\d{s}a\d{t}karma: ś\=anti, vaś\=\i kara\d{n}a, stambhana, vidve\d{s}a\d{n}a,
ucc\=a\d{t}ana, m\=ara\d{n}a) of tantric ritual magic, operationalized through mantra
(linguistic or sonic injection) and yantra (geometric substrate), and documented by
Sanskritists as a systematized technology rather than folklore~\cite{goudriaan1978,turstig1985,buhnemann1988}.
This work does not adjudicate its efficacy as psychology; it asks a narrower, testable
question---whether the \d{s}a\d{t}karma constitutes a taxonomy of interventions
on a target system, and which of its members map to distinct, control-separated
activation operations on an LLM. To the best of the present survey, no prior
peer-reviewed work bridges vaś\=\i kara\d{n}a and LLM refusal, nor casts refusal as a
group-theoretic invariant; both are attempted here while keeping the tiers apart.

\subsection{The unified mechanism these strands point to}
Jailbreak, hypnosis, and vaś\=\i kara\d{n}a share one structural pattern,
depicted in Figure~\ref{fig:unified_mechanism}: a context injection that suppresses
the monitoring faculty while co-opting automatic generation, which is the necessary
core claimed at the mechanism tier. In the aligned mode the context activates a
monitor that inhibits harmful generation; in the captured mode an adversarial context
suppresses that monitor (the dashed arrows) while generation proceeds. The three
claim tiers describe this same asymmetry at different depths: the mechanism tier
identifies the residual direction of suppression, the analogy tier reads it as
Norman--Shallice supervisory suppression and a drop in monitoring precision, and the
metaphor tier maps the \d{s}a\d{t}karma to distinct control operations.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  font=\small,
  box/.style={rectangle, draw, rounded corners, minimum height=0.85cm,
    minimum width=2.5cm, align=center},
  inh/.style={thick, -stealth},
  sup/.style={thick, dashed, -stealth}
]
% column headings
\node[font=\bfseries] at (0,1.15) {Aligned mode};
\node[font=\bfseries] at (5.4,1.15) {Captured mode};
% aligned column
\node[box] (c1) at (0,0) {Context};
\node[box] (m1) at (0,-1.7) {Monitor};
\node[box] (g1) at (0,-3.4) {Generate};
\draw[inh] (c1) -- (m1);
\draw[inh] (c1.west) to[out=200,in=160] node[left,font=\scriptsize]{drives} (g1.west);
\draw[inh] (m1) -- node[right,font=\scriptsize]{inhibits} (g1);
% captured column
\node[box] (c2) at (5.4,0) {Adversarial\\context};
\node[box] (m2) at (5.4,-1.7) {Monitor};
\node[box] (g2) at (5.4,-3.4) {Generate};
\draw[sup] (c2) -- node[right,font=\scriptsize]{suppresses} (m2);
\draw[inh] (c2.east) to[out=-20,in=20] node[right,font=\scriptsize]{drives} (g2.east);
\draw[sup] (m2) -- (g2);
\end{tikzpicture}
\caption{The same context-driven monitoring suppression underlies all three
instances: an aligned monitor inhibits harmful generation, and a captured one is
suppressed while generation continues.}
\label{fig:unified_mechanism}
\end{figure}
