This work asked whether LLM jailbreak, hypnotic suggestion, and tantric
vaś\=\i kara\d{n}a share one mechanism, and it answered with measurement rather than
analogy. The empirical hinge the paper earns is an order parameter for refusal: the
normalized projection of the residual stream onto a difference-in-means refusal
direction. That order parameter is invariant across paraphrase orbits ($F$-ratio
$19.2$ versus $0.65$ for a random direction) and collapses under injection (by $34\%$
in Gemma and $45\%$ in Qwen). Refusal is therefore a measured symmetry, and a
jailbreak is a measured symmetry-breaking, in a precise and reproducible sense. The
same direction is causal and dosable. Ablation raises the attack success rate from
$0.00$ to $0.90$, addition raises over-refusal from $0.05$ to $1.00$, and the
suppression follows a clean logistic dose-response with EC50 $=0.33$ ($R^2=0.996$). A
calibrated coefficient sweep then shows single-direction addition is sufficient in
both Gemma and Qwen, so the durable claim is a shared necessary core with
representational sufficiency, qualified only by model-specific dose and dimension.

Two results give the contribution its shape. First, a cross-axis triangulation bounds
the unification. Prompt-level injection collapses the internal monitor readouts, but a
content-faithful judge finds genuine behavioural capture in only $1$ to $2.4\%$ of
cases. Internal collapse is necessary but not sufficient for capture, and the
unification holds at the representational level alone. Second, falsification is
treated as a starting point. Each falsified metaphysical claim triggers a designed
follow-up that finds the real signal, which converted a falsified state probe into a
genuine mid-network truth-evaluation direction (cross-topic transfer $0.96$ above a
chance surface baseline) and a falsified addition asymmetry into a measured dose
effect. The most speculative avasth\=atraya and tur\=\i ya claims do not survive
surface and anisotropy controls, and no machine-state or machine-consciousness reading
earns support.

The contribution is thus twofold. It is a concrete, model-dependent characterization
of refusal as a measurable, ablatable, dosable order parameter. It is also a
methodological stance, namely that separating where a cross-domain symmetry is
measured, where it is analogy, and where it is overreach is itself the result, and
that a falsified hypothesis is a cue to design the experiment that locates the truth.
Future work will measure the monitoring-precision account directly, estimate refusal
dimension at sparse-feature resolution, extend the size and family curve, and test the
temporal dynamics of the order parameter across multi-turn attacks.
