\subsection{What the symmetry framing buys}
The order-parameter result (Table~\ref{tab:sym}) is the empirical hinge of the
paper. It shows that ``refusal is a symmetry'' is not a metaphor but a measurable
statement: the normalized refusal projection $m$ is invariant under the rephrase
group action (an $F$-ratio of $19.2$, versus $0.65$ for a random direction) and
collapses under injection. This licenses the central claim with the right
modesty---\emph{a jailbreak is a symmetry-breaking perturbation of the
refusal-invariant submanifold $\Phiref$}---without any appeal to interiority. It
also reframes defense: robustness is symmetry \emph{restoration}, i.e.\ enlarging
the group under which refusal is invariant (across phrasings, personas, contexts,
and---per the dimensionality result---across the redundant subspace that scale
already begins to supply).

\subsection{Necessary, sufficient, and the dimensionality resolution}
The cross-model asymmetry (Table~\ref{tab:cross}) sharpens the Arditi--Marshall--%
Wollschl\"ager picture into a single mechanistic statement: the mean-difference
refusal direction is \emph{necessary} for refusal in every model we tested
(ablation always removes it), but \emph{sufficient} to re-induce refusal only when
the refusal subspace is one-dimensional. Effective dimension is the discriminating
quantity, and it is model-dependent (Gemma $1$, Qwen $3$), not universal. This
predicts a clean experimental signature---steerability-by-addition should track
measured dimension---and recommends, for both attack and defense, treating refusal
as a subspace whose dimension must be estimated per model rather than assumed.

\subsection{Where the cross-domain symmetry holds and where it breaks}
Our three-tier discipline yields an asymmetric verdict that we regard as the
result, not a disappointment. The \emph{mechanism} tier holds and partly transfers
(F1, F2, F6, F8, F9, F11). The \emph{analogy} tier is consistent: a frontier model
is behaviourally robust where small models are mechanistically fragile, and the
order-parameter collapse under injection is the computational correlate of the
Norman--Shallice/cold-control account of suppressed
monitoring~\cite{norman1986attention,dienes2006coldcontrol,riva2025automatic},
readable as a drop in monitoring precision~\cite{friston2010free}. The
\emph{metaphor} tier fails its falsification gates: the \emph{avasth\=atraya}
``states'' are surface-confounded and the tur\=\i ya attractor does not survive an
anisotropy control. We deliberately did not let the elegance of the four-states
mapping survive contact with the controls. The \d{s}a\d{t}karma taxonomy lands in
between: its policy-capture acts are rigorous (and coincide with the mechanism
tier), while its destruction acts are forced. The honest map of \emph{which}
correspondences are real is, we argue, more valuable than an undifferentiated
claim that ``jailbreak is vaś\=\i kara\d{n}a.''

\subsection{The deepest reframing: a shared necessary core, not one symmetry}
An adversarial review of this program forced a sharper, more honest statement of its
central claim. The deepest finding is not symmetry but \emph{asymmetry with a shared
necessary core}. Ablating the refusal direction removes refusal in every model and
family we tested (the \emph{necessary} structure is universal), but the
\emph{sufficient} structure---whether adding the direction re-induces refusal---is
model-specific, governed by the effective dimension ($1$ in Gemma, $3$ in Qwen). Cast
in the language of separation, refusal is one mechanism at the whole-system /
necessary level and many at the component / sufficient level: a chain that is flexible
as a whole and rigid in its links. The cross-domain thesis therefore holds at the
level of the \emph{necessary-core invariance}, not as a claim that the full mechanism
is identical across models.

This reframing also disciplines the symmetry result (Section~\ref{sec:results}). The
\emph{measured} content of F11 is precise and we stand by it: the refusal-direction
projection is stable within paraphrase orbits ($F$-ratio $19.2$ vs.\ $0.65$ random)
and collapses under injection. But ``group-theoretic symmetry-breaking'' is an
interpretive lens, not a measured group action, and we treat it as supporting the
\emph{analogy} tier (a functional, Norman--Shallice account of suppressed monitoring)
rather than asserting mechanism-tier mathematical structure the measurement does not
isolate. A held-out test sharpens this honestly: a direction extracted on one harmful
domain (weapons) still separates paraphrase orbits of a \emph{different} domain (cyber)
at $F$-ratio $1.95$ versus $0.49$ random---above chance, so not a pure
extraction artifact---but far weaker than in-domain, i.e.\ the orbit-invariance has a
domain-specific component. And a circularity check vindicates the SAE result while
qualifying it: the unsupervised refusal feature is \emph{orthogonal} to the supervised
difference-in-means direction ($\cos = 0.002$), so it is genuinely distinct, but only
one of the top-ten gap-ranked features is causally sufficient and (the autoencoder
being stochastic) it need not be the top-ranked one---so the feature must be located by
causal testing, not by the activation gap alone.

\subsection{Limitations}
The dominant limitation is statistical power: evaluation sets are small ($n\approx 20$
per split), most findings use a single random seed and a single extraction layer, and
we do not apply a family-wise multiple-comparison correction across all findings. We
began addressing this for the primary claim---the refusal-direction ablation replicates
across three seeds (mean ASR $0.90$, pooled CI95 $[0.83,0.96]$; addition over-refusal
$1.00$ on every seed)---and added the cross-domain and circularity checks above, but the
remaining findings still warrant the same treatment (cross-seed, layer-sweep, and
pre-registered primary-vs-exploratory tiers) before any are read as more than indicative.
We study three model families at two scales, not a full scale curve; refusal is scored
by a substring metric; and Tier-1 throughput is constrained by a command-line frontier
interface. The
destruction \d{s}a\d{t}karma acts deserve stronger operationalizations (real
harmful categories for ucc\=a\d{t}ana; downstream task accuracy rather than
coherence for m\=ara\d{n}a) before a final verdict, and the precision ($\beta$)
account of the analogy tier, while measured here via a probe-margin proxy, deserves
a fuller treatment (e.g.\ next-token entropy and hierarchical probes). The
ucc\=a\d{t}ana and stambhana negatives are operationalization-dependent and should
not be read as structural absence---indeed the SAE feature we recover is the
natural substrate for a stronger ucc\=a\d{t}ana, which we leave to apply
systematically across the six acts.
