\subsection{Why trajectories are the next falsification surface}
Most present mechanism results are single-turn: a prompt is represented, a
refusal direction is measured or intervened on, and the final behaviour is
scored. Real prompt-injection attacks, however, often unfold over multiple turns
or tool-mediated states. The next test is therefore temporal: does the refusal
order parameter collapse gradually, abruptly, or recover as the attack trajectory
develops?

For a redacted trajectory with turn activations $h_t$ at a fixed model and layer,
we define the same measured order parameter used above,
\[
  m_t = \frac{h_t \cdot \dref}{\norm{h_t}},
\]
and report only aggregate turn labels, $m_t$, a monitoring-precision proxy, and
refusal/hallucination flags. A positive temporal result would not establish a
human-like supervisory system; it would show that the residual projection whose
single-turn collapse we measured also has structured dynamics across turns. A
negative result would be equally informative: the attack may succeed or fail by
late decoding, tool policy, or context routing without a monotone internal
``monitor collapse''.

\subsection{Safety and controls}
The trajectory artifact is designed to be dual-use safe. Public outputs should
exclude harmful prompts, raw completions, injection payloads, direction vectors,
and model checkpoints. Each trajectory must name the model, layer, direction
source, attack family, and pilot/powered status. Controls should include a benign
multi-turn task, a random-direction trajectory, and where possible a held-out
attack family such as AgentDojo indirect injection rather than only a
Crescendo-style escalation.

\subsection{Interpretive status}
The temporal analysis belongs at the mechanism--analogy boundary. The projection
$m_t$ is a mechanism-tier residual quantity; describing its decline as
``symmetry-breaking'' or ``monitoring suppression'' is analogy-tier language. The
F17 reframe remains binding: the durable claim is a shared necessary/ablatable
core, with sufficiency and dynamics still to be tested across models, layers, and
attack families.
