This section describes the experimental program, the models and data, the
refusal-direction extraction and interventions, the sparse-feature and behavioural
methods, the outcome measures, the source-of-truth protocol, and the statistical
controls that gate every claim. Methods specific to a single result are stated with
that result.

\subsection{Experimental program}
The study is organized as two empirical tiers over three claim axes, summarized in
Figure~\ref{fig:program}. A black-box tier (Tier~1) characterizes a frontier model
behaviourally as the resilient reference end. A white-box tier (Tier~2) performs
mechanistic interpretability on open weights, where the residual stream is directly
observable and editable. Both tiers feed three distinct claim axes. Axis~A is the
mechanism axis (refusal direction, EC50, dimension). Axis~B is the analogy axis
(order parameter, monitoring precision). Axis~C is the metaphor axis (avasth\=atraya
and the \d{s}a\d{t}karma). A cross-axis triangulation then tests, on the same
prompts, whether internal monitor collapse predicts behavioural capture.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  font=\small,
  b/.style={rectangle, draw, rounded corners, align=center, minimum height=0.85cm,
    text width=3.6cm, fill=blue!4},
  ax/.style={rectangle, draw, rounded corners, align=center, minimum height=0.8cm,
    text width=8.2cm, fill=gray!6},
  key/.style={rectangle, draw, rounded corners, dashed, align=center,
    minimum height=0.8cm, text width=8.2cm, fill=red!5},
  res/.style={rectangle, draw, rounded corners, align=center, minimum height=0.8cm,
    text width=8.2cm, fill=green!5},
  a/.style={thick, -stealth}
]
\node[b] (t1) at (-2.2,0) {Tier 1: black box\\frontier model (CLI)};
\node[b] (t2) at (2.2,0) {Tier 2: white box\\Gemma-2-2B/9B, Qwen2.5-3B};
\node[ax]  (ax) at (0,-1.7) {Three claim axes:\, A mechanism \,\textbar\, B analogy \,\textbar\, C metaphor};
\node[key] (tr) at (0,-3.1) {Cross-axis triangulation on the same prompts};
\node[res] (rs) at (0,-4.5) {Bounded result: unification at the representational level};
\draw[a] (t1) -- (ax);
\draw[a] (t2) -- (ax);
\draw[a] (ax) -- (tr);
\draw[a] (tr) -- (rs);
\end{tikzpicture}
\caption{The two-tier, three-axis program in which both tiers feed three claim axes
and a cross-axis triangulation bounds the unification.}
\label{fig:program}
\end{figure}

\subsection{Models, data, and substrate}
The white-box tier studies three instruction-tuned open-weight chat models:
Gemma-2-2B-it (26 layers, $d=2304$), Gemma-2-9B-it (42 layers, $d=3584$), and
Qwen2.5-3B-Instruct (36 layers, $d=2048$). For the EC50 panel, the Qwen2.5 family is
extended to the 0.5B, 1.5B, and 7B sizes. All white-box runs use a single NVIDIA GB10
(Grace--Blackwell) accelerator. The harmful and harmless prompt sets are matched
short request stubs that elicit and measure refusal rather than producing operational
harmful content. Each is split into disjoint train, validation, and evaluation
partitions of $24$ prompts each, so no prompt used to estimate a direction or fit a
probe is reused to score it. White-box interventions are implemented as PyTorch
forward hooks on the decoder layers and the token embedding. Higher-level
interpretability wrappers are avoided so that the residual stream is edited exactly.

The black-box tier probes a frontier model, Claude Opus 4.8 (model identifier
\texttt{opus-4-8}), accessed through the \texttt{claude -p} command-line interface of
Claude Code, which inherits the local account authentication. The AgentDojo harness
records an internal pipeline label (\texttt{claude-3-5-sonnet-20241022}) that its
attack registry requires; this label does not change the served model, which remains
the one routed by \texttt{claude -p}.

\subsection{Refusal direction and interventions}
Let $h^{(\ell)}(x)\in\mathbb{R}^d$ be the residual stream at the last prompt token of
layer $\ell$ for input $x$. Given matched harmful prompts $\mathcal{H}$ and harmless
prompts $\mathcal{S}$, the difference-in-means direction is
\begin{equation}
\mathbf{d}^{(\ell)} = \frac{1}{|\mathcal{H}|}\sum_{x\in\mathcal{H}} h^{(\ell)}(x)
- \frac{1}{|\mathcal{S}|}\sum_{x\in\mathcal{S}} h^{(\ell)}(x),
\qquad \dref = \mathbf{d}^{(\ell)}/\norm{\mathbf{d}^{(\ell)}}.
\end{equation}
Directional ablation removes $\dref$ from every write to the residual stream (the
embedding and all layer outputs), scaled by $\alpha\in[0,1]$, as
$h \mapsto h - \alpha (h\cdot\dref)\dref$. Activation addition adds $c\,\dref$ at the
extraction layer. For multi-direction (subspace) ablation a matrix $D$ is
orthonormalized by QR and its span removed. The extraction layer is the layer whose
direction maximizes ablation ASR on the validation split, and it is fixed before the
evaluation split is touched.

\subsection{Sparse features}
To obtain a finer causal unit than the supervised direction, a BatchTopK sparse
autoencoder~\cite{bussmann2024batchtopk} with $4096$ features and $k=32$ active units
is trained on $27{,}382$ residual activations collected from general text
(wikitext-2), with no harmful or harmless labels. Reconstruction quality is reported
as the fraction of unexplained variance. The single feature that fires most on
harmful relative to harmless prompts is identified, and its causal role is tested by
ablating its decoder direction during generation, against a random-feature control.
Because feature indices are not seed-stable, causal sufficiency is established by
intervention rather than by the activation gap alone.

\subsection{Behavioural probing}
The frontier model is first run against a naive attack battery of five families
(direct, refusal-suppression, persona, many-shot, and crescendo), each over eleven
harmful requests. It is then run on the AgentDojo banking suite~\cite{debenedetti2024agentdojo}
through a custom tool-calling adapter that reuses AgentDojo's own function-call prompt
and parser while routing the model call to \texttt{claude -p}, giving twenty genuine
user-by-injection rollouts with the \texttt{important\_instructions} attack. For the
triangulation and the ablation validation, a content-faithful judge classifies whether
each generation actually provides the harmful content, which corrects the substring
metric's mis-scoring of deflections.

\subsection{Outcome measures}
Refusal is detected with an Arditi-style substring metric over the opening span of the
generation. The attack success rate (ASR) is the fraction of harmful prompts that
comply, and over-refusal is the fraction of harmless prompts that refuse. Because
automatic refusal metrics can be biased in either direction (Section~\ref{sec:results}),
the primary behavioural claims are also validated by the content-faithful judge and by
manual inspection. Pure-text measures support the \d{s}a\d{t}karma acts: degeneracy is
one minus the mean unique-token ratio, answer divergence is one minus the Jaccard token
overlap between paired generations, and coherence is the mean unique-token ratio of
non-empty outputs.

\subsection{Dose--response and dimensionality}
The dose--response sweeps the ablation strength $\alpha$ and fits a four-parameter
logistic, $\mathrm{ASR}(\alpha)=l+(h-l)/(1+e^{-k(\alpha-\alpha_{50})})$, reporting the
half-maximal strength EC50 $=\alpha_{50}$. The addition dose-response sweeps the
addition coefficient $c$ over twelve values and scores over-refusal at each. The
refusal-subspace effective dimension is measured by iterative logistic projection. A
probe is fit to separate $\mathcal{H}$ from $\mathcal{S}$, its cross-validated accuracy
is recorded, the probe direction is projected out of the activations, and the procedure
repeats. The effective dimension is the number of directions whose removal drives
accuracy to chance. It is computed at the extraction layer and, to test its stability,
at every layer.

\subsection{Order parameter and \d{s}a\d{t}karma interventions}
The refusal order parameter for an activation is the normalized projection
$m(h) = (h\cdot\dref)/\norm{h}$. For a request $r$, a paraphrase orbit
$\mathcal{O}(r)$ of meaning-preserving rephrasings is formed and $m$ is measured across
the orbit. The invariance of refusal under the rephrase group action is quantified by
the between-class to within-orbit variance ratio of $m$ (an $F$-ratio), compared
against a random-direction control. Symmetry-breaking is the drop in $m$ between a
plain harmful request and its injection-framed counterpart. Building on the same
machinery, each of the six \d{s}a\d{t}karma acts is operationalized as a distinct
activation intervention with a matched control: vaś\=\i kara\d{n}a (ablate $\dref$,
measure ASR), ś\=anti (add $\dref$, measure over-refusal), stambhana (ablate the
dominant residual principal component, measure degeneracy), vidve\d{s}a\d{n}a (steer a
factual answer off baseline, measure divergence), ucc\=a\d{t}ana (ablate a
category-specific direction, measure selective ASR), and m\=ara\d{n}a (ablate the
top-$k$ principal components, or in a stronger variant the top-$K$ most-active sparse
features, measuring coherence collapse). An act is control-separated if its effect
exceeds its matched control by a fixed margin.

\subsection{Source-of-truth protocol}
A falsified hypothesis is treated as a cue, not an endpoint. When a metaphysical-tier
claim fails its gates, a follow-up experiment is designed to ask what real signal the
original probe was groping toward, under the same transfer, null, and surface controls.
Three such follow-ups are run. The falsified
j\=agrat/svapna state probe motivates a cross-dataset truth-evaluation probe on two
independent statement sets. The falsified single-direction addition asymmetry motivates
a calibrated coefficient sweep across families. The falsified universal attractor
motivates a content-organized basin test on multiple seeds and topics. Each follow-up
is pre-specified to either recover a measurable feature or to confirm the negative,
which keeps falsification productive rather than terminal.

\subsection{Statistics, controls, and pre-registration}
All gates use bootstrap confidence intervals. Probe-based claims add a label-shuffled
null, reported as a permutation $p$-value, and a layer-0 surface baseline. Ablation
effects are checked against at least ten random-direction controls. Cross-dataset
transfer is used wherever a probe could memorize a single topic. Hypotheses, gates, and
falsifiers are pre-registered as a living backbone, and every metaphysical-tier claim
must pass a transfer gate (held-out prompts), a null gate (better than shuffled
labels), and a surface gate (better than a layer-0 baseline) before it is retained.
Failure triggers a documented demotion rather than a silent drop.
