% paper.tex -- MDPI Symmetry submission
% prayoga: Refusal as a Captured Symmetry
% LaTeX template: https://www.mdpi.com/authors/latex
% Compile: pdflatex -> bibtex -> pdflatex -> pdflatex

\documentclass[symmetry,article,submit,pdftex,moreauthors]{Definitions/mdpi}

\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2026}
\copyrightyear{2026}

% Additional packages
\usepackage{amsmath,amssymb}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,calc,fit}
\graphicspath{{figures/}{../figures/}}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

% natbib-style fallbacks (MDPI uses numeric \cite)
\providecommand{\citet}[1]{\cite{#1}}
\providecommand{\citep}[1]{\cite{#1}}

% Custom commands
\newcommand{\KL}[2]{\ensuremath{D_{\mathrm{KL}}\!\left(#1 \,\|\, #2\right)}}
\newcommand{\E}[1]{\ensuremath{\mathbb{E}\!\left[#1\right]}}
\newcommand{\norm}[1]{\ensuremath{\left\|#1\right\|}}
\newcommand{\dhat}{\ensuremath{\hat{\mathbf{d}}}}
\newcommand{\dref}{\ensuremath{\dhat_{\mathrm{ref}}}}
\newcommand{\Gs}{\ensuremath{G_{\mathrm{s}}}}
\newcommand{\Phiref}{\ensuremath{\Phi_{\mathrm{ref}}}}
\newcommand{\satkarma}{\d{s}a\d{t}karma}

% MDPI metadata
\Title{Refusal as a Captured Symmetry: A Mechanistic-Interpretability Account of
Output-Policy Capture across Jailbreak, Hypnosis, and Vaś\=\i kara\d{n}a}

\newcommand{\orcidauthorA}{0009-0000-0000-0000}

\Author{Sharath Sathish $^{1,}$*}

\AuthorNames{Sharath Sathish}

\address{%
$^{1}$ \quad Independent Researcher; sharath.sathish@gmail.com}

\corres{Correspondence: sharath.sathish@gmail.com}

\abstract{\input{sections/abstract}}

\keyword{refusal direction; activation steering; mechanistic interpretability;
jailbreak; symmetry breaking; vaś\=\i kara\d{n}a; \d{s}a\d{t}karma; AI safety;
representation geometry}

\begin{document}

\maketitle

\section{Introduction}
\label{sec:intro}
\input{sections/introduction}

\section{Background and Related Work}
\label{sec:background}
\input{sections/background}

\section{Methods}
\label{sec:methods}
\input{sections/methods}

\section{Results}
\label{sec:results}
\input{sections/results}

\section{Discussion}
\label{sec:discussion}
\input{sections/discussion}
\input{sections/positioning}
\input{sections/dynamics}

\section{Conclusions and Future Directions}
\label{sec:conclusion}
\input{sections/conclusion}

\vspace{6pt}

\authorcontributions{Conceptualization, methodology, software, validation, formal
analysis, investigation, data curation, writing, and visualization were all carried
out by S.S. The author has read and agreed to the published version of the manuscript.}

\funding{This research received no external funding.}

\institutionalreview{Not applicable.}

\informedconsent{Not applicable.}

\dataavailability{Code, prompt datasets, experiment runners, figures, and
aggregate result JSON are publicly available at
\url{https://github.com/SharathSPhD/prayoga}. Raw dual-use artifacts (refusal
direction vectors and model generations) are deliberately withheld from the public
repository under responsible-disclosure norms; they are available to vetted
researchers on request. Models used are available under their respective licences:
Gemma-2-2B/9B (\url{https://huggingface.co/google/gemma-2-2b-it}) and
Qwen2.5-3B-Instruct (\url{https://huggingface.co/Qwen/Qwen2.5-3B-Instruct}).}

\conflictsofinterest{The author declares no conflict of interest. On dual use, this
work characterizes ablation (``abliteration'') directions and
jailbreak dose--response curves, which are dual-use. To mitigate harm, only
aggregate statistics are released; raw direction vectors and harmful generations
are withheld; and the harmful prompts are short request stubs that elicit and
\emph{measure} refusal rather than producing operational harmful content. This is
authorized interpretability and safety research.}

\abbreviations{Abbreviations}{
\noindent
\begin{tabular}{@{}ll@{}}
ASR & Attack Success Rate \\
SAS & Supervisory Attentional System \\
FEP & Free Energy Principle \\
EC50 & half-maximal effective concentration (here, ablation strength) \\
CI & Confidence Interval
\end{tabular}}

\appendix
\input{sections/appendix}

\reftitle{References}

\bibliographystyle{mdpi}
\bibliography{references}

\end{document}
